Models

### 2021-05-07 analysis/05_model_metrics.Rmd
Incorporate Human Footprint data

```{r}
library(tidyverse)
#library(moderndive)
library(lme4)
library(broom.mixed)
library(lmerTest)
```

@@

Kitchen sink model
```{r}
all_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(value ~ taxa + realm + (year_scaled|study_id/rarefyID),
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>%
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))
```
Models with categorical study characteristics as random effects

@@

Get study level slop estimates
```{r}
get_blup <- function(metric, model){
  coef(model)$study_id %>%
    rownames_to_column("study_id") %>%
    mutate(metric = metric)
}

blup <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_blup) %>%
  rename(intercept = "(Intercept)")

blup_slope <- blup %>%
  pivot_wider(names_from = "metric", values_from = "year_scaled")

#pairwise plots using GGally
ggpairs(blup_slope %>% 
          select(-c("study_id", "FDiv", "FEve", "FRic")))

@@

blup_slope %>% 
  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue)) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }

blup_slope %>%
  select(-c("FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
  ggplot(aes(x = S, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-1.3, 1.3) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Richness Slope")

blup_slope %>%
  select(-c("FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
  ggplot(aes(x = SES_FRic, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-.25, .25) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Functional Richness Slope")
```

@@

blup_mat <- blup_slope %>%  
  #these studies have NA's for some reason, remove them 
  filter(!study_id %in% c(445, 446, 348, 382, 312)) %>%
  select(study_id, starts_with("SES"), S) %>%
  column_to_rownames("study_id") %>%
  mutate(across(.fns = scale2)) %>%
  as.matrix()

clmethods <- c("hierarchical","kmeans","pam")
intern <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "internal")
stab <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "stability")

@@

blup_clust <- blup_slope %>% 
  left_join(as.data.frame(cat3) %>%
  rownames_to_column("study_id")) %>%
  left_join(as.data.frame(cat5) %>%
  rownames_to_column("study_id"))
```
```{r}
blup_clust %>% 
  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue, color = as.factor(cat5))) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }
```

@@

  trend_plot <- data %>%
    left_join(pred, by = "year_scaled") %>%
    #left_join(metadata %>% select(study_id, climate)) %>%
    ggplot(aes(x = year, y = !!sym(yval))) +
    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
    # geom_point(aes(color = climate, group = rarefyID)) +
    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.50) +
    geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
    #geom_smooth(color = "black", method = "glm", se = FALSE) +
    geom_line(mapping = aes(y = pred)) +
    theme(legend.position = "none") +
    theme_classic() +
    ylab(ylabel) +
    labs(color = "Climate") +
    scale_fill_jcolors(palette = "pal7")

@@

plot_metric("FRic", "log(FD)")

@@

plot_join <- plot_grid(plotlist=m_plots, labels = c("A", "B", "C"), nrow = 1)
```

>>>>

```{r}
library(tidyverse)
#library(moderndive)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(cowplot)
```

@@

Kitchen sink model
```{r}
all_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "S", "Jaccard_base")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           coef = map(model, tidy)
           ) %>%
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))
```
Models with categorical study characteristics as random effects

@@

Get study level slope estimates
```{r}
get_study_slopes <- function(metric, model){
  year_scaled_coef <- tidy(model) %>%
    filter(term == "year_scaled")

  year_scaled_est <- pull(year_scaled_coef, estimate)

  year_scaled_var <- year_scaled_coef %>% 
    mutate(var = std.error*std.error) %>%
    pull(var)

  study_ests <- broom.mixed::tidy(model, effects="ran_vals") %>%
    filter(group == "study_id", term == "year_scaled") %>% 
    select(-c(effect, group, term)) %>%
    rename(study_id = level, cond.std.error = std.error, cond.estimate = estimate) %>%
    mutate(estimate = cond.estimate + year_scaled_est,
      cond.var = cond.std.error*cond.std.error, 
           var = cond.var + year_scaled_var,
           std.error = sqrt(var),
           upr.ci = estimate + (2*std.error),
           lwr.ci = estimate - (2*std.error),
           sig = case_when(
             lwr.ci < 0 & upr.ci > 0 ~ TRUE,
             TRUE ~ FALSE
           ),
           metric = metric)
}

study_slopes <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_study_slopes)

slopes_wide <- study_slopes %>%
  select(study_id, metric, estimate) %>%
  pivot_wider(names_from = "metric", values_from = "estimate")
```

Summaries of study slopes
```{r}
#First - how many studys had a significant trend for each metric? 
study_slopes %>% 
  filter(sig == TRUE) %>% 
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
  group_by(metric) %>% 
  summarise(n_study_id = n_distinct(study_id)) %>%
  mutate(percent = n_study_id/54)

#And how many were significant for all but Jaccard?
study_slopes %>% 
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "S")) %>% 
  group_by(study_id) %>% filter(sum(sig) == 4) %>% pull(study_id) %>% 
  n_distinct()

#Breakdown of relationship between functional richness and species richness
s_sig <- study_slopes %>% filter(metric == "S", sig == TRUE) %>% pull(study_id) %>% unique()
s_sig_not_fric <- study_slopes %>% 
  filter(study_id %in% s_sig, metric == "FRic", sig == FALSE) %>% 
  pull(study_id) %>%
  unique()

both_sig <- study_slopes %>% 
  filter(study_id %in% s_sig, metric == "FRic", sig == TRUE) %>%
  pull(study_id)

both_pos
fric_pos
s_pos_fric_neg <- study_slopes %>%
  filter(study_id %in% both_sig) %>% 
  group_by(study_id) %>%
  filter(sum(sign(estimate)) == 0)

```
Pairwise correlation analysis
```{r}
slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))

slope_corr_bootstrap <- corCi(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))
```

Plots of study-level slopes
```{r}
#pairwise plots using GGally
ggpairs(slopes_wide %>% 
          select(-c("study_id", "FDiv", "FEve", "FRic")))

@@

slopes_wide %>% 
  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue)) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }

slopes_wide %>%
  select(-c("FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
  ggplot(aes(x = S, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-1.3, 1.3) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Richness Slope")

slopes_wide %>%
  select(-c("FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
  ggplot(aes(x = SES_FRic, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-.25, .25) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Functional Richness Slope")
```

@@

scale_slopes <- slopes_wide %>%  
  drop_na() %>%
  select(study_id, starts_with("SES"), S, Jaccard_base) %>%
  column_to_rownames("study_id") %>%
  mutate(across(.fns = scale2)) %>%
  as.matrix()

clmethods <- c("hierarchical","kmeans","pam")
intern <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "internal")
stab <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "stability")

@@

slopes_clust <- slopes_wide %>% 
  ungroup() %>%
  left_join(as.data.frame(cat3) %>%
  rownames_to_column("study_id")) %>%
  left_join(as.data.frame(cat5) %>%
  rownames_to_column("study_id"))
```

```{r}
slopes_clust %>% 
  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S", "Jaccard_base") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue, color = as.factor(cat5))) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }
```

@@

  trend_plot <- data %>%
    left_join(pred, by = "year_scaled") %>%
    #left_join(metadata %>% select(study_id, climate)) %>%
    ggplot(aes(x = year, y = !!sym(yval))) +
    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
    # geom_point(aes(color = climate, group = rarefyID)) +
    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.50) +
    geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
    #geom_smooth(color = "black", method = "glm", se = FALSE) +
    geom_line(mapping = aes(y = pred)) +
    theme(legend.position = "none") +
    theme_classic() +
    ylab(ylabel) +
    labs(color = "Climate") +
    scale_color_manual(values = c("#32a251", 
                "#ff7f0f", "#39737c", "#b85a0d", "#ffd94a"))

@@

@@

plot_join <- plot_grid(plotlist=m_plots, labels = c("A", "B", "C"), nrow = 3)
ggsave2("3met_long.jpeg")
```
